Comparing software variability across and within fMRI analysis packages

### 2021-12-31 paper/paper.tex
Update results at t=17bits

\maketitle
\begin{abstract}
  Variability has been broadly observed in functional MRI analyses as a result
  of software differences
  both between and within analytic tools. However, the
  relationship between within-tool and between-tool software variabilities for a
  given analysis is unclear. We extended a previous comparison of fMRI
  analysis software libraries (namely FSL, AFNI, and SPM) and related the
  observed differences to within-tool software variability
  simulated using Monte-Carlo arithmetic. In group analyses, we found that
  between-tool software variability was consistently larger than within-tool software
  variability simulated at the magnitude of
  machine error. In subject analyses, the same within-tool software variability
  approached between-tool software variability in some regions for some subjects.
  Interestingly, within- and between-tool software variabilities appeared moderately
  correlated ($r > 0.5$) in all tested conditions. Finally, we found that
  within-tool software variability had effects of similar magnitude to between-tool
  software variability when simulated from numerical perturbations introduced at the virtual
  precision of 17 bits --- a high but plausible level of uncertainty for
  scientific computing libraries. \tristan{Revise from latest results}
  %  \gk{What does this mean? They only have
  % precision of 17, or they use data that is 17 bits?}
  %  \yohan{agree with Greg, it is unclear and a strong assertion!}
  %  \camille{It is not entirely clear to me how this relates to the previous
  %  conclusions that BT variability was consistently larger than machine
  %  error} \tristan{Better?}
   Our findings motivate the
  continued investigation of within-tool software variability in neuroimaging, and position it
  as a possible proxy for studying some of the arbitrary between-tool software variations.
\end{abstract}

@@

\subsection{At precision t=17~bits and in FSL, WT approached BT in the group analysis}
\tristan{We cannot say that from the data that we have.}
\tristan{Discuss subject-level BA plots.}
While the previous results were obtained by simulating WT with numerical perturbations 
at the magnitude of machine error, we also
evaluated numerical variability across different virtual precisions for FSL
and found that the virtual precision of t=17 bits minimized the RMSE
between BT and WT in unthresholded group maps.
FSL produced unthresholded group maps with $\mu=0.483$ and $\sigma=0.410$ at this virtual precision,
which was still lower than BT variability
(Wilcoxon signed-rank test and t-test p~\textless~$10^{-5}$ for all tests).
However, Figure~\ref{fig:gnp-mni} shows that
BT and WT reached comparable magnitudes in some regions. BT and WT remained
moderately correlated at this precision (Pearson's r
in [0.53, 0.61], p$< 10^{-5}$, Figure~\ref{fig:gnp-mni}-\textbf{B}).

%%%%%%%%%% plot at precision t=17%%%%%%%%
\begin{figure*}[ht]
  \begin{subfigure}[ht]{\textwidth}
    \centering
    \includegraphics[width=.55\textwidth]{figures/p17-diffs.pdf}
    % \includegraphics[width=.75\textwidth]{figures/bg_global_precision.pdf}
    %\caption{Standard deviation of thresholded t-statistics map on template surface}
  \end{subfigure}
  %\caption{Standard deviation of thresholded t-statistics map on template surface}
  \caption{Unthresholded group t-statistics differences computed
  within tools at the virtual precision of t=17~bits (\textbf{A}),
  and voxel-wise comparison of absolute differences between BT and WT (\textbf{B}).}
  \label{fig:gnp-mni}
\end{figure*}

@@

%%%%%%%%%% Var. of Thresh %%%%%%%%
\begin{figure*}[ht]
  \begin{subfigure}[ht]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/act_deact-marked.pdf}
    % \includegraphics[width=\textwidth]{figures/plots/wt-tstats/act_deact-framed.png}
    % \includegraphics[width=\textwidth]{figures/abs/gl-thresh.png}
    %\caption{Standard deviation of thresholded t-statistics map on template surface}
  \end{subfigure}
  \centering
  \caption{\textbf{A},\textbf{B},\textbf{C}: Thresholded group-level
    t-statistics within tools at machine error for FSL (\textbf{A}), SPM
    (\textbf{B}) and AFNI (\textbf{C}). Arrows point to activation clusters
    impacted by both within- and between-tool variability \tristan{Ali: is it really the case? Can you make the arrows point more accurately to 
    the clusters? Some of them are quite far from the cluster they point to.}.
     \textbf{D}: Confusion matrices of
    activation instability in BT and WT among the 360 regions of the
    HCP-MMP1.0 parcellation.}
  \label{fig:thresh-maps}
\end{figure*}

@@

\tristan{This paragraph needs to be revised as we can't conclude that from the data we have}
The finding that numerical variability approached tool variability at the
virtual precision of t=17~bits is interesting too. Indeed, while machine
error generally introduces perturbations on the least-significant bit of the mantissa,
ones 
% \yohan{I would not mix virtual precision and error measurement}\tristan{better?}
common scientific software dependencies introduce larger perturbations
.
For instance, SciPy's 2D spline interpolation was recently found to be
precise up to 10~bits~\cite{pytracer}
and its replacement by a more precise implementation might therefore introduce
numerical perturbations leading to errors in the range of between-tool
variability. Such high errors
might be triggered by updates in operating systems, Python, MATLAB, and
other software dependencies.
% \yohan{The sentence is ok, just a picky point: the SciPy result you cited
% describes the precision of the spline implementation while at the end
% you're talking differences due to updates. The nuance is that two versions
% may have a similar precision and so when you udpate your system, the
% difference between both versions is low but each version can be
% ill-conditionned.}
% \tristan{better?}

@@

Figure~\ref{fig:replication-diff} shows the difference in SPM and AFNI
group analyses maps between the results in~\cite{bowring2019exploring} and
our replication. Numerical perturbations of 1 ulp are likely to have been
introduced by our replication due to the use of GNU/Octave vs MATLAB for SPM
and multithreading for AFNI.
 \camille{Would this be feasible to check?}
 \tristan{Ali, could you answer that?}
  However, the group maps remained very similar
overall, which led us to conclude that our results correctly reproduced
the ones in~\cite{bowring2019exploring}.
\begin{figure*}[ht]
  % \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
  % \centering
  \includegraphics[width=\textwidth]{figures/replication-diffs.png}
  \caption{Differences between reproduced and original results obtained in~\cite{bowring2019exploring}
    of unthresholded group-level t-statistics for SPM (left) and AFNI
    (right). The highest areas of difference in AFNI seem to be due to
    differences in brain masks.}
  \label{fig:replication-diff}
  % \end{minipage}}
\end{figure*}

>>>>

\maketitle
\begin{abstract}
  Variability has been broadly observed in functional MRI analyses as a result
  of software differences
  both between and within analytic tools. However, the
  relationship between within-tool and between-tool software variabilities for a
  given analysis is unclear. We extended a previous comparison of fMRI
  analysis software libraries (namely FSL, AFNI, and SPM) and related the
  observed differences to within-tool software variability
  simulated using Monte-Carlo arithmetic. In group analyses, we found that
  between-tool software variability was consistently larger than within-tool software
  variability simulated at the magnitude of
  machine error. In subject analyses, the same within-tool software variability
  approached between-tool software variability in some regions for some subjects.
  Interestingly, within- and between-tool software variabilities appeared moderately
  correlated ($r > 0.5$) in all tested conditions. 
  % Finally, we found that
  % within-tool software variability had effects of similar magnitude to between-tool
  % software variability when simulated from numerical perturbations introduced at the virtual
  % precision of 17 bits --- a high but plausible level of uncertainty for
  % scientific computing libraries. \tristan{Revise from latest results}
  % %  \gk{What does this mean? They only have
  % precision of 17, or they use data that is 17 bits?}
  %  \yohan{agree with Greg, it is unclear and a strong assertion!}
  %  \camille{It is not entirely clear to me how this relates to the previous
  %  conclusions that BT variability was consistently larger than machine
  %  error} \tristan{Better?}
   Our findings motivate the
  continued investigation of within-tool software variability in neuroimaging, and position it
  as a possible proxy for studying some of the arbitrary between-tool software variations.
\end{abstract}

@@

% \subsection{At precision t=17~bits and in FSL, WT approached BT in the group analysis}
% \tristan{We cannot say that from the data that we have.}
% \tristan{Discuss subject-level BA plots.}
% While the previous results were obtained by simulating WT with numerical perturbations 
% at the magnitude of machine error, we also
% evaluated numerical variability across different virtual precisions for FSL
% and found that the virtual precision of t=17 bits minimized the RMSE
% between BT and WT in unthresholded group maps.
% FSL produced unthresholded group maps with $\mu=0.483$ and $\sigma=0.410$ at this virtual precision,
% which was still lower than BT variability
% (Wilcoxon signed-rank test and t-test p~\textless~$10^{-5}$ for all tests).
% However, Figure~\ref{fig:gnp-mni} shows that
% BT and WT reached comparable magnitudes in some regions. BT and WT remained
% moderately correlated at this precision (Pearson's r
% in [0.53, 0.61], p$< 10^{-5}$, Figure~\ref{fig:gnp-mni}-\textbf{B}).

% %%%%%%%%%% plot at precision t=17%%%%%%%%
% \begin{figure*}[ht]
%   \begin{subfigure}[ht]{\textwidth}
%     \centering
%     \includegraphics[width=.55\textwidth]{figures/p17-diffs.pdf}
%     % \includegraphics[width=.75\textwidth]{figures/bg_global_precision.pdf}
%     %\caption{Standard deviation of thresholded t-statistics map on template surface}
%   \end{subfigure}
%   %\caption{Standard deviation of thresholded t-statistics map on template surface}
%   \caption{Unthresholded group t-statistics differences computed
%   within tools at the virtual precision of t=17~bits (\textbf{A}),
%   and voxel-wise comparison of absolute differences between BT and WT (\textbf{B}).}
%   \label{fig:gnp-mni}
% \end{figure*}

@@

%%%%%%%%%% Var. of Thresh %%%%%%%%
\begin{figure*}[ht]
  \begin{subfigure}[ht]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{figures/act_deact-marked.pdf}
    % \includegraphics[width=\textwidth]{figures/plots/wt-tstats/act_deact-framed.png}
    % \includegraphics[width=\textwidth]{figures/abs/gl-thresh.png}
    %\caption{Standard deviation of thresholded t-statistics map on template surface}
  \end{subfigure}
  \centering
  \caption{\textbf{A},\textbf{B},\textbf{C}: Thresholded group-level
    t-statistics within tools at machine error for FSL (\textbf{A}), SPM
    (\textbf{B}) and AFNI (\textbf{C}). Arrows point to activation clusters
    impacted by both within- and between-tool variability.
     \textbf{D}: Confusion matrices of
    activation instability in BT and WT among the 360 regions of the
    HCP-MMP1.0 parcellation.}
  \label{fig:thresh-maps}
\end{figure*}

@@

% \tristan{This paragraph needs to be revised as we can't conclude that from the data we have}
% The finding that numerical variability approached tool variability at the
% virtual precision of t=17~bits is interesting too. Indeed, while machine
% error generally introduces perturbations on the least-significant bit of the mantissa,
% ones 
% % \yohan{I would not mix virtual precision and error measurement}\tristan{better?}
% common scientific software dependencies introduce larger perturbations
% .
% For instance, SciPy's 2D spline interpolation was recently found to be
% precise up to 10~bits~\cite{pytracer}
% and its replacement by a more precise implementation might therefore introduce
% numerical perturbations leading to errors in the range of between-tool
% variability. Such high errors
% might be triggered by updates in operating systems, Python, MATLAB, and
% other software dependencies.
% \yohan{The sentence is ok, just a picky point: the SciPy result you cited
% describes the precision of the spline implementation while at the end
% you're talking differences due to updates. The nuance is that two versions
% may have a similar precision and so when you udpate your system, the
% difference between both versions is low but each version can be
% ill-conditionned.}
% \tristan{better?}

@@

Figure~\ref{fig:replication-diff} shows the difference in SPM and AFNI
group analyses maps between the results in~\cite{bowring2019exploring} and
our replication. Numerical perturbations of 1 ulp are likely to have been
introduced by our replication due to the use of GNU/Octave vs MATLAB for SPM
and multithreading for AFNI.
%  \camille{Would this be feasible to check?}
%  \tristan{Ali, could you answer that?}
%  \ali{Yes, all the output data are available in Beluga cluster in `~/projects/def-glatard/between-tool-samples/results/ds000001/'.
%  These replicated results can find in particular in `tools/AFNI/replicated' for AFNI and `tools/SPM/Octave' for SPM results.}
  However, the group maps remained very similar
overall, which led us to conclude that our results correctly reproduced
the ones in~\cite{bowring2019exploring}.
\begin{figure*}[ht]
  % \fbox{\begin{minipage}{\dimexpr \textwidth-2\fboxsep-2\fboxrule}
  % \centering
  \includegraphics[width=\textwidth]{figures/replication-diffs.png}
  \caption{Differences between reproduced and original results obtained in~\cite{bowring2019exploring}
    of unthresholded group-level t-statistics for SPM (left) and AFNI
    (right). The highest areas of difference in AFNI seem to be due to
    differences in brain masks.}
  \label{fig:replication-diff}
  % \end{minipage}}
\end{figure*}
